Significant declines in standardised test scores due to COVID-19 school closures disproportionately affect vulnerable students

### 2024-02-18 paper/paper.qmd
Fixed Typos

Finally, data from “Pandemic	Schooling	Mode	and	Student	Test	Scores:	Evidence	from	US	School Districts” [@Jack_Oster_2] was used to examine changes in district-level pass rate data from 21 state standardised assessments, taken in the spring of 2018, 2019 and 2021. Both @fig-mathscoresbyonline and @fig-elascoresbyonline both provide information on the change in pass rates in Math and ELA standardised tests from the previous year, respectively, taking into account the share of in-person learning done by various school districts and across the three years of interest, each indicated by its respective color. Consistent with Jack and Oster [-@Jack_Oster_2023], our analysis shows a significant decline in pass rates from the 2018-2019 to the 2020-2021 school year. We see that in Spring 2021, there is a clear fall in the pass rates overall in both subject areas from 2019 (note that there were no standardised assessmments in 2020), though this fall is much more pronounced in Math than ELA (-12.8 percentage points compared to the -6.8 percentage points, respectively). The discrepancy between these two subjects is consistent with much of the current economic literature, which shows mathematics scores to be more sensitive to schooling differences [@betts2011effect; @angrist].

@@

In line with some of our previous discussions, we looked towards building upoing Jack and Oster's work by examining the potential systematic disparities between relative student racial compositions and the average percentage point change in test pass rates. In particular, both @fig-mathscoresrace and @fig-elascoresrace provide information on the changes in test pass rates across the three years of interest from the year prior as a function of the race quantiles that each schooling district belonged to for each of the share of White, Black, and Hispanic students. By grouping the test score data by state then calculating the average share of each of the three races mentioned, we were able to assign each schooling district a value of "High" or "Low" for each of the three races, detailing whether the particular schooling district had a share of students of a particular race greater than, or lower than, the median share across the 21 states, respectively. The figures distinguish between the year and race in question by shape and color, respectively.

@@

(1) Between January 2020 and December 2021, the United States experienced the second-highest number of days of school closed worldwide, with an average of 667 days. In fact, it falls second only to Azerjiban, which recorded an average of 669 days.

@@

Looking at @fig-race alongside @fig-mathscoresrace and @fig-elascoresrace provides insights into the impacts of virtual schooling. As noted by Jack and Oster [-@Jack_Oster_2023], it can be seen that the equity impacts are two-fold. First, as shown in @fig-race, schools (and students) of more vulnerable racial backgrounds -- especially, Black and Hispanic -- are less likely to have access to in-person instruction. Second, as shown in @fig-mathscoresrace and @fig-elascoresrace, for such groups, the consequences of a lack of in-person instruction on academic performance are more severe. We can see how the the districts with a higher share of black and Hispanic students had more virtual days, and also had large decrease in their test scores relative to the districts with lower shares. On the other hand, the districts with a high share of white students had less virtual school days, and had a smaller decrease in their pass rates relative to the districts with lower shares.

>>>>

Finally, data from “Pandemic	Schooling	Mode	and	Student	Test	Scores:	Evidence	from	US	School Districts” [@Jack_Oster_2] was used to examine changes in district-level pass rate data from 21 state standardised assessments, taken in the spring of 2018, 2019 and 2021. Both @fig-mathscoresbyonline and @fig-elascoresbyonline both provide information on the change in pass rates in Math and ELA standardised tests from the previous year, respectively, taking into account the share of in-person learning done by various school districts and across the three years of interest, each indicated by its respective color. Consistent with Jack and Oster [-@Jack_Oster_2023], our analysis shows a significant decline in pass rates from the 2018-2019 to the 2020-2021 school year. We see that in Spring 2021, there is a clear fall in the pass rates overall in both subject areas from 2019 (note that there were no standardised assessments in 2020), though this fall is much more pronounced in Math than ELA (-12.8 percentage points compared to the -6.8 percentage points, respectively). The discrepancy between these two subjects is consistent with much of the current economic literature, which shows mathematics scores to be more sensitive to schooling differences [@betts2011effect; @angrist].

@@

In line with some of our previous discussions, we looked towards building upon Jack and Oster's work by examining the potential systematic disparities between relative student racial compositions and the average percentage point change in test pass rates. In particular, both @fig-mathscoresrace and @fig-elascoresrace provide information on the changes in test pass rates across the three years of interest from the year prior as a function of the race quantiles that each schooling district belonged to for each of the share of White, Black, and Hispanic students. By grouping the test score data by state then calculating the average share of each of the three races mentioned, we were able to assign each schooling district a value of "High" or "Low" for each of the three races, detailing whether the particular schooling district had a share of students of a particular race greater than, or lower than, the median share across the 21 states, respectively. The figures distinguish between the year and race in question by shape and color, respectively.

@@

(1) Between January 2020 and December 2021, the United States experienced the second-highest number of days of school closed worldwide, with an average of 667 days. In fact, it falls second only to Azerbaijan, which recorded an average of 669 days.

@@

Looking at @fig-race alongside @fig-mathscoresrace and @fig-elascoresrace provides insights into the impacts of virtual schooling. As noted by Jack and Oster [-@Jack_Oster_2023], it can be seen that the equity impacts are two-fold. First, as shown in @fig-race, schools (and students) of more vulnerable racial backgrounds -- especially, Black and Hispanic -- are less likely to have access to in-person instruction. Second, as shown in @fig-mathscoresrace and @fig-elascoresrace, for such groups, the consequences of a lack of in-person instruction on academic performance are more severe. We can see how the districts with a higher share of black and Hispanic students had more virtual days, and also had large decrease in their test scores relative to the districts with lower shares. On the other hand, the districts with a high share of white students had less virtual school days, and had a smaller decrease in their pass rates relative to the districts with lower shares.
